---
title: "2022 GDJC Talks"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    self_contained: false
---

# June 2022

### Dr. Lauren Wedekind - Investigating the relationship between birth weight and type 2 diabetes

<iframe width="560" height="315" src="https://www.youtube.com/embed/XTW3ZAz2EuA" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


Our June GDJC Talk, marking 3 years since our first one, was given by Lauren Wedekind, one of our GDJC organizers. Lauren shared work from her PhD research on diabetes and birth weight and summarized some previous work that inspired it. Themes she reviewed included the Developmental Origins of Health and Disease (DOHaD) hypothesis, the foetal insulin hypothesis, and the genetic and non-genetic influences on the relationship between birth weight and diabetes (which in some study populations has been found to be U-shaped). She summarized her work on genome-wide association studies of birth weight and using clinical and genetic information to predict incident diabetes: click [here](https://scholar.google.com/citations?user=fsXJ8RkAAAAJ&hl=en) for more articles and abstracts she has co-authored on these topics.

Lauren used this talk as a practice presentation for her PhD thesis defense, which was closed to the public, so she invited friends, family and colleagues. Thus, this talk was specifically designed to be accessible to all, and used terms that those without formal training in diabetes or epidemiology can better understand. If you are interested in giving a practice presentation with us through our GDJC Talks platform, click on our ["Want to give a talk?" tab](https://gdjc.github.io/website/want_to_give_a_talk.html).

<<<<<<< HEAD
```{r june-2022, echo=FALSE}

library("vembedr")

embed_url("https://youtu.be/XTW3ZAz2EuA")

```
=======
>>>>>>> ba7421780f65ab5c0026546563de2f7fde0b3ff4

# May 2022

### Dr. Gabriela Carrillo Balam - Sex differences in diabetes and depression

```{r may-2022, echo=FALSE}

embed_url("https://youtu.be/4plc-A0eMaA")

```

Dr. Carrillo gave a talk about her ongoing study showing some interesting differences in the prevalence of diabetes and depression in Mexican men and women. Note, the study is not yet published, meaning that the results have yet to be peer reviewed. We will give an update, when the study has been published. 

The study was based on data from the Mexican National Health and Nutrition Survey 2018-2019 and the study protocol was pre-registered [here](https://osf.io/g3bsk)

In the talk she describes the overlapping etiology of diabetes and depression within biological, environmental, social and behavioral aspects. 

The key findings in the study was that women had a higher prevalence of diabetes than men and women also had a higher prevalence of depression than men. In particular, their interaction analysis suggested that there was an additive interaction. Meaning that women with diabetes was more likely to have depression above what is expected based on having diabetes or being a women alone. 

# April 2022

### Dr. Cassie Mitchell and Statistician Emma Stinson - Metabolic responses to a mixed meal tolerance test in individuals at risk of type 2 diabetes

```{r april-2022, echo=FALSE}

embed_url("https://youtu.be/6Zp7l9Ayv94")

```

[Dr. Cassie Mitchell, PhD, RD (postdoctoral fellow)](https://www.researchgate.net/profile/Cassie-Mitchell) and [Emma Stinson, MPH (statistician)](https://www.researchgate.net/profile/Emma-Stinson) at [NIH/NIDDK Phoenix Epidemiology and Clinical Research Branch](https://www.niddk.nih.gov/research-funding/at-niddk/labs-branches/phoenix-epidemiology-clinical-research-branch) jointly presented their work on metabolic responses to a mixed meal tolerance test in individuals at risk of type 2 diabetes. 

In their study, they compared the association between different metrics, such as the area under the curve, from the oral glucose tolerance test or a mixed meal tolerance test and risk of type 2 diabetes. 

In the Q&A there were several interesting questions. Firstly, it is worth noting that the population in this study were relatively young (mean age about 26 years) and were all mostly indigenous populations which are known to have a higher genetic predisposition to develop type 2 diabetes. This may make it hard to compare the results to other populations. Overall, the oral glucose tolerance test provided greater absolute glucose values due to a higher dose compared to the mixed meal. The mixed meal may not stress the system in a similar way and hence may not be able to detect early stages of risk. 

On the other hand, a mixed meal challenge may be easier to handle for the participants. The mixed meal used here consisted of 40 energy from carbohydrates, 40% from fat and 20% from protein. The total energy contribution was 33% of total daily energy need. One issue with the mixed meal is, however, that it is hard to standardize across studies because there is no consensus on the composition or which foods to use. 

One of the advantages of using a mixed meal challenge is that is may reveal other metabolic dysfunctions than only glucose which the oral glucose tolerance test is designed for. However, more research from this project will cast light on this question. 

Lastly, future investigations of this dataset will look at various other measures of the glucose curves to examine other differences between the mixed meal challenge and oral glucose tolerance test.

# March 2022

### Anthony Muchai Manyara - Comparison of risk factors between people with and without type 2 diabetes in Nairobi

```{r march-2022, echo=FALSE}

embed_url("https://youtu.be/s1j_b-yfUBM")

```

Dr. Anthony Muchai Manyara from the University of Glasgow, gave our March 2022 GDJC Talk on the topic of risk factors in people with and without type 2 diabetes in Nairobi, Africa. The talk was based on [this paper](https://onlinelibrary.wiley.com/doi/10.1111/tmi.13629). 

The discussion after the talk focused on how the controls were sampled and potential issues with this. Although the methods for doing this in the study was not ideal (asking participants to bring a frend of similar age and sex), there was also a practical consideration that this was deemed the best way of recruiting controls under the circumstances. 

The study also revealed some interesting findings regarding differences in grip strengths and central adiposity between those with and without type 2 diabetes.

All in all, the study points to some risk factors like grip strength and central adiposity as being important to monitor in this setting prospective studies are needed to investigate these further.

# February 2022

### Research Associate, Alisa Kjaergaard - Mendelian randomization: causal inference without intervention?

```{r february-2022, echo=FALSE}

library("vembedr")

embed_url("https://youtu.be/_lZdkg_yNqw")

```

Global Diabetes Journal Club Talk by Research Associate Alisa Kjaergaard from the Steno Diabetes Center Aarhus, Aarhus, Denmark, on Mendelian randomization with examples related to diabetes research. Dr. Kjaergaard gave an introduction to the key concepts of Mendelian Randomization. Indeed, this talk is very relevant to those new to the topic of Mendelian Randomization.

Her most recent published studies include Mendelian Randomization and regular observational studies investigating the link between [obesity and kidney funtion](https://academic.oup.com/clinchem/advance-article-abstract/doi/10.1093/clinchem/hvab249/6469631?redirectedFrom=fulltext&login=false), [thyroid function and sex hormones](https://link.springer.com/article/10.1007%2Fs10654-021-00721-z) and [homocysteine and female fertility](https://www.nature.com/articles/s41430-021-00898-2).

During the Q&A, which was not recorded, we also the use of causal language, the issue with pleiotropy in two-sample Mendelian randomization and issues with using machine learning methods to identify a large number of genetic polymorphisms associated with an exposure of interest.  

For more information about Dr. Kjaergaard follow this [link](https://www.stenoaarhus.dk/kontakt/Alisa-Kjaergaard/)
